A Mechanised Library for Generic (Co)Programming

### 2024-01-21 paper/paper.tex
small edits

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example of this fact is the quicksort algorithm given by
the \haskell{partition} and \haskell{quicksort} function below (written in
Haskell code):
\begin{minted}{haskell}
partition :: [a] -> ([a], a, [a])
partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)

@@

Since then this result has had huge impact in the functional programming
community for structuring functional programs in a way that it guarantees
termination (or productivity) in programming languages with general
recursion~\cite{DanielssonHJG06}.

There is a tension:
\begin{itemize}
  \item On one hand we can allow general recursion and the employ recursion
schemes to help the programmer structuring programs so that they terminate
  \item on the other hand we could forbid non-termination, but at this point
there is no algorithm to tell termination and non-termination apart.
\end{itemize}

The first option, helps the programmer but does not guarantee termination, since
the programmer can still feed to the recursion scheme a functor that does not
have a fixed-point or that has an algebra that does not terminate.

The second option forbids non-termination, but does not allow all terminating
programs.

With this work we intend to begin a study of the applications of recursion
schemes that goes beyond functional programming, that is to dependent type
theories such as a the ones employed in proof assistants like Agda or Coq where
non-termination is outright rejected.  This has been a long-standing challenge
to come up with a sound algorithm which rejects all programs that do not
terminate (soundness) and accepts as many well-defined programs as possible
(completeness). Our aim is to provide an additional tool to the every-day
programmer to be able to extend the range of recursive programs we can write.

>>>>

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorithm. This is given by the \haskell{partition} and \haskell{quicksort} function below (written in
Haskell code):
\begin{minted}{haskell}
partition :: [a] -> ([a], a, [a])
partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)

@@

This result has given a much needed unifying view of recursion scheme which
helped understanding their nature and also in providing a unified way of
reasoning about recursive programs by means of fusion laws.

All in all, the functional programming way of dealing with recursion is to just
allow all types of recursion and then restricting ourselves to only the
recursion schemes provided by the given libraries. This gives a way of ``fast
and loose reasoning''~\cite{DanielssonHJG06} which works provided that the user
is somewhat aware of the perils of feeding this libraries with potentially
non-terminating programs.

The type theoretical approach instead is to reject all definitions which cannot
be proven well-defined. It is safer, but incomplete.

With this work we intend to begin a study of the applications of recursion
schemes that goes beyond functional programming, that is to dependent type
theories such as a the ones employed in proof assistants like Agda or Coq where
non-termination is outright rejected.

This has been a long-standing challenge to come up with a sound algorithm which
rejects all programs that do not terminate (soundness) and accepts as many
well-defined programs as possible (completeness). Our aim is to provide an
additional tool to the every-day programmer to be able to extend the range of
recursive programs we can write.
